Ornaments as indicators of Social change before and after European contact at Kiwulan, Northeastern Taiwan

### 2019-08-01 analysis/paper/paper.Rmd
update

The direct impacts of European colonialism on indigenous communities in East Asia were much less pronounced than island Southeast Asia and Oceania. Throughout East Asia direct colonial rule was rare and limited, but the question of long-lasting indirect impacts on local indigenous communities remains largely unanswered. Understanding the indirect effects of colonialism are important for detecting colonial impacts on Indigenous peoples in the periphery of colonial control [@TorrenceandClarke2000; @Trabert2017]. In many parts of the world, the introduction of foreign trade goods by colonial traders into local Indigenous societies caused significant transformations of Indigenous economic, cultural, and socio-political systems [@Dietler1997; @Junker1993; @Mitchell2000; @Silliman2005]. The agency of Indigenous peoples in the colonial contexts has been broadly discussed to understand the negotiations between colonized and colonizer, and the resistance and accommodations of Indigenous people through their daily cultural practices [@Dietler2015; @Given2004; @Rubertone2000; @Silliman2001; @Torrence2000; @TorrenceandClarke2000; @Voss2005]. Studying the consumption of foreign goods is one way to understand Indigenous experiences of colonial encounters and explore the agency of Indigenous peoples by examining how goods were distributed and used [@Dietler2005; @Mullins2011; @Scaramelli2005].

Taiwan provides an ideal case study of indirect effects of colonialism because it has archaeological evidence spanning its initial involvement in regional exchange networks of East Asia and Southeast Asia since the tenth century. It became part of global trade systems during the early 17th century due to western European expansions that connected Europe and Asia [@Liu2017]. Despite colonization of Taiwan by the Spanish and Dutch from 1624 to 1662, there has been little discussion of the archaeology of Indigenous responses to the European colonization. @Berrocal2018 study \the long-term record of Heping Dao\ in northern Taiwan and explore the effects of European colonization on the local populations by focusing on the environmental change and introduction of diseases, with a little mention of indigenous social complexity. Yilan in Northeastern Taiwan is an especially unique location for exploring indirect effects due to its physically isolated location. The rugged \XXX mountains to the west of the Yilan Plain that made it not easily accessible to Europeans and the colonial control was weaker compared to other parts of this region [@Andrade2007; @Kang2012]. The sea provides a way through which Indigenous people in Yilan could interact with other groups from different regions, including other native Indigenous groups, foreign traders such as the Han people of China, and Europeans. Large-scale and frequent trade activities brought a wide variety of trade objects into Indigenous communities in Yilan through local regional exchange network. One of the most common types of trade goods in this region were personal ornaments such as glass and stone beads was one of the common foreign goods found at the settlement sites in this region [@Chen2007; @Li2014; @NMTH2005]. Personal adornments are a social signal of an individual’s status [@Joyce2005; @Scaramelli2005]. For example, the consumption of stone beads in Southeast Asia during Iron Age is often associated with increasing social stratification or socio-political complexity [@Bellina2014; @Carter2016; @Francis2002; @Theunissen2000].

In this paper, I explore the evidence from Yilan spanning from late Iron Age to historical period to address the question of whether indirect colonial influences on Indigenous populations can be detected. I focus on  the ornaments recovered from the archaeological excavations at Kiwulan (1400-1900 AD), a large Iron Age settlement in northeastern Taiwan [@Chen2007]. Prior to the contact with the Europeans in the early 17th century, northern Taiwan had been involved in both regional and long-distance East and Southeast Asian trade systems, and \later encountered a large wave of Han Chinese migration in the 19th century [@Chen2005]. Northeastern Taiwan provides a good example to explore the indirect colonial impacts on Indigenous societies by comparing archaeological evidence from different episodes of culture contact [\cf.@Lape2003]. I will describe the variety and cultural context of trade ornaments in this region, and examine their potential of the ornaments to inform on social changes relating to trade with colonial groups. I hypothesize that there was a greater diversity of ornaments types and materials at Kiwulan after European contact due to participation in a larger scale exchange network. I also predict that after European contact there will be new spatial patterns in the distributions of ornaments at Kiwulan that indicates increasing social inequality. I hypothesize that after Chinese contact there was a decline in the production, use and discard of ornaments at Kiwulan due to \???.

@@

Prior to European colonization, the indigenous people in northern Taiwan had already established their own exchange network that has been described as "inter-insular trade" by @Chen2005 [pp. 12] to refer to small-scale regional trade between China and Taiwan on irregular basis. This local exchange network of northeastern Taiwan was increasingly influenced by long distance trade between Fuzhou or Quanzhou in China and Ryukyu in Japan since 15th century due to its location on the shipping routes. This was intensified in the 17th century with the arrival of the Europeans and their trade goods [@Chen2005; @Liu2017; @Wang2007]. Due to its location on the periphery of major trade ports, \Keelung and \Tamsui, in northern Taiwan, Yilan was involved in this network through the periodic exchange activities between different indigenous groups. Yilan, also called "Kavalan" by local indigenous people, is an alluvial plain circumscribed by the Pacific and mountains on the sides. Rivers and seas provide the way for the interaction between local indigenous settlements and further \afield communities, such as the Basai in northern Taiwan and Han Chinese. Kavalan people offered rice, deer hides, and gold in exchange for beads, metal tools or ornaments, ironwares, porcelains, and textiles with outside traders [@Chen2005; @Hsieh2009; @LiandWu2006].

Historical documents from the Spanish and Dutch in the early 17th century provide some information about the social life and culture of indigenous people in Northern Taiwan when the Europeans arrived. The Spanish founded Fort San Salvador at Keelung in 1626 and Fort San Domingo in 1629 at Tamsui, and sent missionaries to local indigenous settlements in this region for religious purpose [@Blusse2000, pp. 343]. The report of the Dominican priest Fr. Jacinto Esquivel in 1632 mentioned that the Taparri, an indigenous tribe from northern Taiwan, usually used cuentas (agate beads) in exchange for necessities with other indigenous groups. This form of exchange was widespread and even the Spanish soldiers learned to use agate as bargaining chips for gambling, since agate equaled money [@LiandWu2006, pp. 132-149]. Father Jacinto Esquivel also recorded how Indigenous people viewed agate pieces as prestige goods in their culture. A indigenous man who wanted to get married had to pay agate beads to the parents of his future wife. Also, agate or golden beads could be used as to resolve conflicts in their daily lives [@LiandWu2006, pp. 151]. Other records mention that the female shamans (majuorbol) in the tribe would use agate beads as magical items in ritual practice for body healing[@Borao2009, pp. 122-151]. An Indigenous funeral process was recorded, documenting the use of agate beads in ritual contexts, with more agate beads, pottery and cloths placed into the graves of more influential people to indicate a family’s higher prestige [@LiandWu2006, pp. 153]. Those historical accounts indicate that agate beads or golden beads were already treated as prestige goods before the arrival of the Europeans. These were likely introduced earlier from Chinese traders.

In 1642, the Dutch VOC defeated the Spanish and took over their forts in northern Taiwan. They introduced a feudal system in an attempt to control the Indigenous communities by asking Indigenous leaders to attend annual ceremony for demonstrate their loyalty [@Andrade2007, ch. 9; @Kang2016, ch. 4]. The Dutch provided beads and other goods based on the demands of indigenous people to secure alliances in the annual ceremony or during traveling [@Kang2016, ch. 6]. Thus, I assume that the Dutch feudal system resulted in an increase in the amount and diversity of ornaments in northeastern indigenous communities during this period.

Chinese historical records from the \Qing dynasty contain some notes on the purposes of ornaments from Yilan. According to those records, Indigenous people in Yilan had a custom to wear ornaments in ceremonial contexts to display their wealth and status [@Chen1963, pp. 228, 308; @Ke1993, pp. 11, 126; @Yao1996, pp. 77]. Among those ornaments, golden fish-shaped necklaces made of copper threads had higher value due to its delicacy and more materials invested in production. These were usually possessed by wealthy people. Other people wore agate beads or glass beads on their head or neck to participate in ceremonies. In 1895, at the beginning of Japanese colonization, @Ino2002 conducted a field survey for several indigenous groups and found that golden necklaces were not used in Yilan at that time, but elderly people still used beads.

@@

Previous researchers divided the Upper Layer Culture component into six layers spanning from the 14th century to the 19th century with a hundred year intervals according to original radiocarbon dates, excavation depth, and types of diagnostic porcelain [@Hsieh2009; @Wang2011]. The layers were assigned numbers from 1 to 6 where L1 corresponds to the youngest layer and L6, the bottom layer, refers to the oldest layer. However, because the specific methods of assigning layers to a time period was not documented in detail for the previous chronology, I reexamined the excavation records to produce a new chronology for the sampled squares. For each samples square I separated excavation units into three episodes of time: pre-European contact period (L5, L6), European contact period (L4), and Chinese contact period (L2 and L1).

I identify the start of European contact at Kiwulan at 1632, when the local villages were attacked by the Spanish who took revenge on an incident happened earlier that year [@Borao2001, pp. 163]. Later in 1647, the Dutch attacked the indigenous villages and forced them to accept colonial rules and economic demands by paying annual tribute [@Andrade2007]. According to Dutch census reports in 1650, Kiwulan was the largest indigenous settlement in the Yilan Plain, with a population of 840 adults [@Nakamura1938, pp. 12]. \Q:The start of European contact Kiwalan is indicated in the archaeological record by the presence of An-ping jars, stonewares that were largely introduced to Taiwan during the early 17th century. I used the presence and absence of An-ping jars, together with the radiocarbon dates, to identify excavation units associated with the pre-European and post-European periods. European colonization ended in 1662 when they were defeated by the kingdom of Tungning found by Koxinga from China. Later in 1683, the Qing dynasty ruled over Taiwan and a large wave of Han Chinese migrated to Yilan during the late 18th century. I recongnise this as the Chinese colonization period, which is archaeological visible at Kiwulan by \Chinese porcelains and bricks that brought by Chinsese communities [@Hsieh2009]. Also, the description of Chinese migration can be identified from Chinese official records [@Chen1963].

>>>>

The direct impacts of European colonialism on indigenous communities in East Asia were much less pronounced than island Southeast Asia and Oceania. Throughout East Asia direct colonial rule was rare and limited, but the question of long-lasting indirect impacts on local indigenous communities remains largely unanswered. Understanding the indirect effects of colonialism are important for detecting colonial impacts on indigenous peoples in the periphery of colonial control [@TorrenceandClarke2000; @Trabert2017]. In many parts of the world, the introduction of foreign trade goods by colonial traders into local indigenous societies caused significant transformations of indigenous economic, cultural, and socio-political systems [@Dietler1997; @Junker1993; @Mitchell2000; @Silliman2005]. The agency of indigenous peoples in the colonial contexts has been broadly discussed to understand the negotiations between colonized and colonizer, and the resistance and accommodations of indigenous people through their daily cultural practices [@Dietler2015; @Given2004; @Rubertone2000; @Silliman2001; @Torrence2000; @TorrenceandClarke2000; @Voss2005]. Studying the consumption of foreign goods is one way to understand indigenous experiences of colonial encounters and explore the agency of indigenous peoples by examining how goods were distributed and used [@Dietler2005; @Mullins2011; @Scaramelli2005].

Taiwan provides an ideal case study of indirect effects of colonialism because it has archaeological evidence spanning its initial involvement in regional exchange networks of East Asia and Southeast Asia since the \tenth century. It became part of global trade systems during the early 17th century due to western European expansions that connected Europe and Asia [@Liu2017]. Despite colonization of Taiwan by the Spanish and Dutch from 1624 to 1662, there has been little discussion of the archaeology of indigenous responses to the European colonization. @Berrocal2018 study \the long-term record of Heping Dao\ in northern Taiwan and explore the effects of European colonization on the local populations by focusing on the environmental change and introduction of diseases, with a little mention of indigenous social complexity. Yilan in Northeastern Taiwan is an especially unique location for exploring indirect effects due to its physically isolated location. The rugged Hsuehshan Range to the west and Central Range to the south of the Yilan Plain that made it not easily accessible to the European colonists and the colonial control was weaker compared to other parts of this region [@Andrade2007; @Kang2012]. The sea provides a way through which indigenous people in Yilan could interact with other groups from different regions, including other native Indigenous groups, foreign traders such as the Han people of China, and Europeans. Large-scale and frequent trade activities brought a wide variety of trade objects into Indigenous communities in Yilan through local regional exchange network. One of the most common types of trade goods in this region were personal ornaments such as glass and stone beads was one of the common foreign goods found at the settlement sites in this region [@Chen2007; @Li2014; @NMTH2005]. Personal adornments are a social signal of an individual’s status [@Joyce2005; @Scaramelli2005]. For example, the consumption of stone beads in Southeast Asia during Iron Age is often associated with increasing social stratification or socio-political complexity [@Bellina2014; @Carter2016; @Francis2002; @Theunissen2000; @Kenoyer2000].

In this paper, I explore the evidence from Yilan spanning from late Iron Age to historical period to address the question of whether indirect colonial influences on Indigenous populations can be detected. I focus on  the ornaments recovered from the archaeological excavations at Kiwulan (1400-1900 AD), a large Iron Age settlement in northeastern Taiwan [@Chen2007]. Prior to the contact with the Europeans in the early 17th century, northern Taiwan had been involved in both regional and long-distance East and Southeast Asian trade systems, and encountered \several waves of Han Chinese migration in the 19th century [@Chen2005]. Northeastern Taiwan provides a good example to explore the indirect colonial impacts on Indigenous societies by comparing archaeological evidence from different episodes of culture contact [cf.@Lape2003]. I will describe the variety and cultural context of trade ornaments in this region, and examine their potential of the ornaments to inform on social changes relating to trade with colonial groups. I hypothesize that there was a greater diversity of ornaments types and materials at Kiwulan after European contact due to participation in a larger scale exchange network. I also predict that after European contact there will be new spatial patterns in the distributions of ornaments at Kiwulan that indicates increasing social inequality. I hypothesize that after Chinese contact there was a decline in the production, use and discard of ornaments at Kiwulan due to overall population declines and acculturation\.

@@

Prior to European colonization, the Indigenous people in northern Taiwan had already established their own exchange network that has been described as “inter-insular trade” by Chen @Chen2005 [pp. 12] to refer to small-scale regional trade between China and Taiwan on irregular basis. This local exchange network of northeastern Taiwan was increasingly influenced by long distance trade between Fuzhou or Quanzhou in China and Ryukyu in Japan since 15th century due to its location on the shipping routes. This was intensified in the 17th century with the arrival of the Europeans and their trade goods [@Chen2005; @Liu2017; @Wang2007]. Due to its location on the periphery of major trade ports, \Keelung and \Tamsui, in northern Taiwan, Yilan was involved in this network through the periodic exchange activities between different indigenous groups. Yilan, also called "Kavalan" by local indigenous people, is an alluvial plain circumscribed by the Pacific and mountains on the sides. Rivers and seas provide the way for the interaction between local indigenous settlements and further afield communities, such as the Basai in northern Taiwan and Han Chinese. Kavalan people offered rice, deer hides, and gold in exchange for beads, metal tools or ornaments, ironwares, porcelains, and textiles with outside traders [@Chen2005; @Hsieh2009; @LiandWu2006].

Historical documents from the Spanish and Dutch in the early 17th century provide some information about the social life and culture of indigenous people in Northern Taiwan when the Europeans arrived. The Spanish founded Fort San Salvador at Keelung in 1626 and Fort San Domingo in 1629 at Tamsui, and sent missionaries to local indigenous settlements in this region for religious purpose [@Blusse2000, pp. 343]. The report of the Dominican priest Fr. Jacinto Esquivel in 1632 mentioned that the Taparri, an indigenous tribe from northern Taiwan, usually used cuentas (agate beads) in exchange for necessities with other indigenous groups. This form of exchange was widespread and even the Spanish soldiers learned to use agate as bargaining chips for gambling [@LiandWu2006, pp. 132-149]. Esquivel also recorded how indigenous people viewed agate pieces as prestige goods in their culture. A indigenous man who wanted to get married had to pay agate beads to the parents of his future wife. Also, agate or golden beads could be used as to resolve conflicts in their daily lives [@LiandWu2006, pp. 151]. Other records mention that the female shamans (majuorbol) in the tribe would use agate beads as magical items in ritual practice for body healing[@Borao2009, pp. 122-151]. An indigenous funeral process was recorded, documenting the use of agate beads in ritual contexts, with more agate beads, pottery and cloths placed into the graves of more influential people to indicate a family’s higher prestige [@LiandWu2006, pp. 153]. Those historical accounts indicate that agate beads or golden beads were already treated as prestige goods before the arrival of the Europeans. These were likely introduced earlier from Chinese traders.

In 1642, the Dutch VOC defeated the Spanish and took over their forts in northern Taiwan. They introduced a feudal system in an attempt to control the indigenous communities by asking indigenous leaders to attend annual ceremony for demonstrate their loyalty [@Andrade2007, ch. 9; @Kang2016, ch. 4]. The Dutch provided beads and other goods based on the demands of Indigenous people to secure alliances in the annual ceremony or during traveling [@Kang2016, ch. 6]. Thus, I assume that the Dutch feudal system resulted in an increase in the amount and diversity of ornaments in northeastern indigenous communities during this period.

Chinese historical records from the year of 1829, 1837, and 1852 during the Qing dynasty [1616-1911] contain some notes on the purposes of ornaments from Yilan [@Chen1963, pp. 228, 308; @Ke1993, pp. 11, 126; @Yao1996, pp. 77]. According to those records, indigenous people in Yilan had a custom to wear ornaments in ceremonial contexts to display their wealth and status. Among those ornaments, golden fish-shaped necklaces made of copper threads had higher value due to its delicacy and more materials invested in production. These were usually possessed by wealthy people. Other people wore agate beads or glass beads on their head or neck to participate in ceremonies. In 1895, at the beginning of Japanese colonization, @Ino2002 conducted a field survey for several Indigenous groups and found that golden necklaces were not used in Yilan at that time, but elderly people still used beads.

@@

Previous researchers divided the Upper Layer Culture component into six layers spanning from the 14th century to the 19th century with a hundred year intervals according to original radiocarbon dates, excavation depth, and types of diagnostic porcelains [@Hsieh2009; @Wang2011]. The layers were assigned numbers from 1 to 6 where L1 corresponds to the youngest layer and L6, the bottom layer, refers to the oldest layer. However, because the specific methods of assigning layers to a time period was not documented in detail for the previous chronology, I reexamined the excavation records to produce a new chronology for the sampled squares. For each sampled square I separated excavation units into three episodes of time: pre-European contact period (L5, L6), European contact period (L4), and Chinese contact period (L2 and L1).

I identify the start of European contact at Kiwulan at 1632, when the local villages were attacked by the Spanish who took revenge on an incident happened earlier that year [@Borao2001, pp. 163]. Later in 1647, the Dutch attacked the indigenous villages and forced them to accept colonial rules and economic demands by paying annual tribute [@Andrade2007]. According to Dutch census reports in 1650, Kiwulan was the largest indigenous settlement in the Yilan Plain, with a population of 840 adults [@Nakamura1938, pp. 12]. The start of European contact in Kiwalan is indicated in the archaeological record by the presence of An-ping jars, stonewares that were largely introduced to Taiwan during the early 17th century. I used the presence and absence of An-ping jars, together with the radiocarbon dates, to identify excavation units associated with the pre-European and post-European periods. European colonization ended in 1662 when they were defeated by the kingdom of Tungning found by Koxinga from China. Later in 1683, the Qing dynasty ruled over Taiwan and a large wave of Han Chinese migrated to Yilan during the late 18th century. I recongnise this as the Chinese colonization period, which is archaeological visible at Kiwulan by \Chinese porcelains and bricks that brought by Chinsese communities [@Hsieh2009]. Also, the description of Chinese migration can be identified from Chinese official records [@Chen1963].
